---
title: "Limitations"
format:
  html:
   toc: true
   toc-depth: 3
   embed-resources: true
  pdf: default
vignette: >
  %\VignetteIndexEntry{limitations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| warning: false
#| message: false
#| include: false
library(scitargets)
```

# Limitations

## Change in scitargets are not detected by targets.

While targets is good to detect change in functions defined in the global environment, it doesn't detect the internal modification of a package like scitargets.

-   what does that mean for a users: if you update scitargets or another package (Seurat, Azimuth) or even a python dependency such as "umap-learn", it is recommanded that you rerun the entire pipeline.

```{r}
#| eval: false
#| include: true
targets::tar_destroy()      # 1. destroy all steps in the pipeline
targets::tar_delete(steps_concerned_1, steps_concerned_2)       # (optional) delete only the concerned steps
targets::tar_make()         # 2. relaunch the pipeline
quarto::quarto_render(".")  # 3. recompile the report
```

