---
title: "Differential expression with scitargets_dea"
format:
  html:
   toc: true
   toc-depth: 3
   embed-resources: true
  pdf: default
vignette: >
  %\VignetteIndexEntry{scitargets_dea}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    eval: false
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(scitargets)
```

::: callout-note
None of the code in this vignette is executed: single-cell differential
expression is far too heavy (Seurat objects, `DESeq2`, `fgsea`, `topGO`, the
organism annotation databases, ...) to run during a package build. Every chunk
is therefore `eval: false` and meant to be read and adapted, not run as-is.
:::

# Overview

The `scitargets_dea` S7 class stores the differential-expression (DE) analysis
of **one group comparison within one cell state**. "Group" is whatever
biological grouping you want to compare (treatment vs control, responder vs
non-responder, ...) and "cell state" is a cluster or an annotated cell type, so
that genes are only compared between cells of the same type.

A single `scitargets_dea` object holds, for that one comparison × cell state:

-   the **single-cell** DE results (`Seurat::FindMarkers`) and/or the
    **pseudobulk** DE results (`DESeq2`),
-   the **GO** enrichment (one or more of BP / CC / MF) and the **GSEA** results
    (MSigDB collections via `fgsea`),
-   the **pseudobulk PCA** and outlier coordinates,
-   the per-level adjusted p-value cutoffs and the parameters used for the run.

A real analysis produces a **list** of these objects (one per comparison × cell
state). That list is what the report generators consume.

The workflow has three steps:

1.  **Enumerate** the comparisons to run with `dea_comparisons()`.
2.  **Run** each comparison with `run_dea()`.
3.  **Get the results out** — accessor tables, plots, the Quarto report
    generators, and the `xlsx` export.

# 1. Enumerate the comparisons

`dea_comparisons()` reads a metadata `data.frame` and returns one record per
(cell state × group pair) to test. It only keeps a group within a cell state
when it has at least `min_cells_per_group` cells.

```{r}
comparisons <- scitargets::dea_comparisons(
  meta = seurat_obj@meta.data,
  group_by = "condition", # metadata column holding the biological groups
  cluster_by = "clusters_0.5", # compare cells WITHIN the same cluster only
  groups = NA, # NA = every pairwise group comparison;
  #   or e.g. c("treated", "control") for one pair
  clusters = NA, # NA = every cluster; or e.g. c("0", "3")
  min_cells_per_group = 3 # skip a (cluster, group) with fewer cells
)
```

The result is a `data.frame`-like object with the parallel vectors `group_by`,
`cluster_by`, `cluster`, `group1` and `group2` — one row per comparison to run.
`group_by` and `cluster_by` **must be different** columns.

# 2. Run the analysis

`run_dea()` runs **one** comparison and returns a single `scitargets_dea`
object. The minimal call:

```{r}
res <- scitargets::run_dea(
  seurat_obj = seurat_obj,
  group_by   = "condition",
  cluster_by = "clusters_0.5",
  cluster    = "0", # the cell state to analyse
  group1     = "treated", # positive log2FC / NES is "up in group1"
  group2     = "control", # reference group
  level      = "both" # "single_cell", "pseudobulk", or "both"
)
```

## Key parameters

`run_dea()` exposes the knobs below. Anything left at its default reproduces a
sensible standard analysis; the pseudobulk-specific arguments are ignored when
`level = "single_cell"`.

### Which analyses to run

-   `level` — `"single_cell"` (`Seurat::FindMarkers`), `"pseudobulk"` (`DESeq2`),
    or `"both"`.
-   `run_go` / `run_gsea` — toggle the GO and GSEA enrichment. GSEA is only
    computed where its signed gene ranking is valid (otherwise a warning is
    emitted and the report omits that level's GSEA tab): for **single cell** only
    when the `FindMarkers` gene universe is unfiltered (`sc_logfc_threshold = 0`,
    `sc_min_pct = 0`, `sc_only_pos = FALSE`); for **pseudobulk** only with
    `pb_test = "Wald"` (genes ranked by the signed `DESeq2` Wald statistic). With
    the defaults, single-cell GSEA is therefore skipped.
-   `go_params` / `gsea_params` — named lists of GO / GSEA options (ontology,
    algorithm, cutoffs, plot size; the GSEA `pathways`, see below).
-   `species` — `"human"` or `"mouse"`. This is the single source of truth for
    the GO annotation database (`org.Hs.eg.db` / `org.Mm.eg.db`) and the GSEA
    `msigdbr` species; it overrides any organism left in `go_params` /
    `gsea_params`.

### Single-cell controls

These map directly onto `Seurat::FindMarkers` and are ignored for the pseudobulk
level.

-   `sc_test_use` — the differential-expression test (`test.use`, default
    `"wilcox"`).
-   `sc_min_pct` / `sc_logfc_threshold` — minimum detection fraction (`min.pct`,
    default 0.1) and minimum |log2FC| (`logfc.threshold`, default 0.25) for a gene
    to be tested.
-   `sc_only_pos` — keep only genes up in `group1` (`only.pos`, default `FALSE`).

For single-cell GSEA, set all three to their unfiltered values
(`sc_logfc_threshold = 0`, `sc_min_pct = 0`, `sc_only_pos = FALSE`) so the full
gene ranking is available.

### Pseudobulk / DESeq2 model

-   `pseudobulk_unit` — metadata column identifying the biological replicate
    (e.g. `"patient_id"`); cells are summed per replicate before `DESeq2`.
-   `min_replicates` (default 3) — minimum replicates per group for the
    pseudobulk comparison to run; `min_cells_per_sample` drops thin replicates.
-   `pb_test` — `"Wald"` or `"LRT"`.
-   `pb_design` / `pb_reduced` — design and (LRT) reduced model formulas given
    as character strings, passed to `DESeq2` via `as.formula`. Any covariate in
    `pb_design` (e.g. `~ batch + group`) is looked up one value per replicate,
    keyed on `pseudobulk_unit` (it must be constant within a replicate).
-   `pb_low_count_filter = c(min_count, min_samples)` — pre-filter keeping genes
    with at least `min_count` counts in at least `min_samples` samples.
-   `pb_lfc_shrink` — when `TRUE`, add a shrunken log2 fold-change (an
    `apeglm` → `ashr` → `normal` cascade; a message reports which was used).
-   `pb_pca`, `pca_n_top_genes`, `pca_outlier_conf` — compute the pseudobulk PCA
    (`DESeq2` variance-stabilising transform) and flag outlier replicates by
    Mahalanobis distance (`mt::pca.outlier`).
-   `pb_remove_outliers` — drop the flagged outlier replicates from the `DESeq2`
    fit (the replicate / min-cell cutoffs are re-checked afterwards).

### Adjusted p-value cutoffs

-   `padj_cutoff_single_cell` / `padj_cutoff_pseudobulk` (each default 0.05) are
    stored on the object and drive, for that level, the volcano cutoff line and
    the GO foreground-gene selection. `padj_cutoff_pseudobulk` is additionally
    passed to `DESeq2::results(alpha = )`, so DESeq2's independent filtering is
    optimised for that FDR threshold.

## Building the GSEA gene sets

GSEA needs a list of pathways. `get_msigdbr_pathways()` builds it from MSigDB and
you pass it through `gsea_params`:

```{r}
gsea_gene_sets <- scitargets::get_msigdbr_pathways(
  collections = c("Hallmark", "GO:BP", "C7:ImmuneSigDB"),
  species = "Homo sapiens"
)
```

## A fuller example

```{r}
res <- scitargets::run_dea(
  seurat_obj = seurat_obj,
  group_by = "condition",
  cluster_by = "clusters_0.5",
  cluster = "0",
  group1 = "treated",
  group2 = "control",
  level = "both",
  run_go = TRUE,
  run_gsea = TRUE,
  go_params = list(
    ontology = "BP", algorithm = "weight01", statistic = "fisher",
    p_adj_cutoff = 0.05, min_signif_genes = 5, plot_number = 20
  ),
  gsea_params = list(pathways = gsea_gene_sets),
  species = "human",
  # single-cell GSEA needs the full gene ranking -> turn off the FindMarkers
  # pre-filters (otherwise only pseudobulk Wald GSEA is computed):
  sc_min_pct = 0, sc_logfc_threshold = 0, sc_only_pos = FALSE,
  pseudobulk_unit = "patient_id",
  min_replicates = 3,
  pb_test = "Wald",
  pb_lfc_shrink = TRUE,
  pb_remove_outliers = TRUE, # drop PCA-flagged outlier replicates
  padj_cutoff_single_cell = 0.05,
  padj_cutoff_pseudobulk = 0.05
)
```

# 3. Get the results out

`run_dea()` returns one object; the helpers below work on a single object.
`is_scitargets_dea(res)` checks the type, and `normalize_dea_list()` coerces a
single object, a list, or a named branch into a plain list of objects.

## Tables

```{r}
# Differential-expression table for a level.
scitargets::markers_table(res, level = "single_cell")
scitargets::markers_table(res, level = "pseudobulk")

# GO enrichment for one direction / ontology / level.
scitargets::go_table(res, direction = "up", ontology = "BP", level = "single_cell")

# GSEA results for one collection.
scitargets::gsea_table(res, collection = "Hallmark", level = "single_cell")
```

## Plots

Every plot is a static `ggplot2` figure by default; pass `interactive = TRUE`
to get a `ggiraph` version with per-element tooltips.

```{r}
# Volcano (the cutoff line uses the level's stored padj cutoff).
scitargets::volcano_plot(res, level = "single_cell")
scitargets::volcano_plot(res, level = "single_cell", interactive = TRUE)

# Pseudobulk PCA, coloured by group and shaped by outlier status.
scitargets::pca_plot(res, level = "pseudobulk")

# GO barplot (one direction / ontology).
scitargets::go_barplot(res, direction = "up", ontology = "BP", level = "single_cell")

# GSEA barplot. By default the bar shows the FDR-adjusted enrichment of each
# gene set (sign(NES) * -log10(adjusted p)). The gene-level ranking fed to GSEA
# is the raw per-gene p-value for single cell (sign(log2FC) * -log10(p)) and the
# signed DESeq2 Wald statistic for pseudobulk.
scitargets::gsea_barplot(res, collection = "Hallmark", level = "single_cell")
```

## Inside a Quarto report

For a website / report, the generators emit the Quarto + knitr text for whole
sections, which you render with `knitr::knit_child()`. `dea_report_lines()`
builds the full block for **one** comparison (summary, volcano, DE table, GO and
GSEA tabsets); the cross-comparison companions take the **list** of comparisons
and are shown once per document.

```{r}
# Helper that knits generated lines into the current document.
render_child <- function(lines) {
  cat(knitr::knit_child(text = lines, quiet = TRUE), sep = "\n")
}

# `comparisons_list` is a list of scitargets_dea objects (see section 4).
# Cross-comparison summaries, shown once:
render_child(scitargets::composition_lines(
  meta_name = "clinical_meta", state_col = "clusters_0.5"
))
render_child(scitargets::dea_upset_lines(
  comparisons_list,
  obj_name = "comparisons_list", level = "single_cell"
))
render_child(scitargets::dea_gsea_yield_lines(
  comparisons_list,
  obj_name = "comparisons_list", level = "single_cell"
))

# One block per comparison:
for (nm in names(comparisons_list)) {
  render_child(scitargets::dea_report_lines(
    comparisons_list[[nm]],
    fig_id = nm,
    levels = "single_cell",
    interactive_plots = c("volcano", "go", "gsea") # or NA for all-static
  ))
}

# Top differentially-expressed genes across all cells (needs the Seurat object):
render_child(scitargets::dea_top_de_lines(
  comparisons_list,
  seurat_obj = seurat_obj, seurat_name = "seurat_obj",
  level = "single_cell", group_by = "clusters_0.5"
))
```

The generated chunks reference variables **by name** (e.g. `obj_name`,
`seurat_name`, `meta_name`), so those variables must be bound in the environment
where the child is knit.

## Export to xlsx

```{r}
scitargets::write_dea_xlsx(res, out_dir = "./out/clinical_markers")
```

# 4. Using it inside a targets pipeline

In a `targets` pipeline the comparisons are enumerated once, then `run_dea()` is
**mapped** over them. Because `scitargets_dea` is an S7 object, two conventions
matter (see also the *Get started* and *Targets factories* vignettes):

-   `format = "rds"` — the default `qs2` format does not support S7 objects.
-   the command returns `structure(list(res), names = ...)` so the branch stores
    the object itself (instead of having `targets` iterate over its fields), and
    the name makes the comparison easy to find in the report.

```{r}
list(
  # `seurat_merged` and `merged_meta` come from upstream targets (see the
  # "Get started" vignette for a minimal multi-run merge).
  tar_target(
    name = comparisons,
    command = scitargets::dea_comparisons(
      meta = merged_meta,
      group_by = "condition",
      cluster_by = "clusters_0.5",
      groups = NA, clusters = NA, min_cells_per_group = 3
    ),
    description = "Group comparisons to run within each cluster"
  ),
  tar_target(
    name = dea_results,
    command = {
      res <- scitargets::run_dea(
        seurat_obj = seurat_merged,
        group_by = comparisons$group_by,
        cluster_by = comparisons$cluster_by,
        cluster = comparisons$cluster,
        group1 = comparisons$group1,
        group2 = comparisons$group2,
        level = "both",
        gsea_params = list(pathways = gsea_gene_sets),
        pseudobulk_unit = "patient_id",
        pb_remove_outliers = TRUE
      )
      structure(
        list(res),
        names = paste0(
          comparisons$group1, "__", comparisons$group2,
          "_", comparisons$cluster
        )
      )
    },
    pattern = map(comparisons),
    deployment = "main", # keep the S7 object on the main process
    format = "rds", # S7 is not supported by qs2
    description = "Differential expression per comparison"
  ),
  tar_target(
    name = dea_xlsx,
    command = scitargets::write_dea_xlsx(dea_results, out_dir = "./out/dea"),
    pattern = map(dea_results),
    format = "file",
    description = "XLSX export of each comparison"
  )
)
```

The mapped `dea_results` target is the list of `scitargets_dea` objects you load
in a report (`targets::tar_read(dea_results)`) and feed to the generators above.
